---
title: "Introduction to Data Science - Workshop Edition"
subtitle: "Cleaning and wrangling data with `janitor` and `forcats`"
author: "Luis Fernando Ramirez, Elena Dreyer and Shruti Kakade"
institute: "Hertie School | [GRAD-C11/E1339](https://github.com/intro-to-data-science-23/workshop-presentations)" #"`r format(Sys.time(), '%d %B %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Welcome to Janitor! 
In this crash course, we will learn how to clean data in R using a library called Janitor. 

```{r, message = F}
#Preparing the libraries
library(readr)
library(tidyverse)
library(janitor)
library(kableExtra)
```
By using a real-life dataset, we want to demonstrate the perks `janitor` has to offer. First we need to download [this dataset](https://datos.cdmx.gob.mx/dataset/interrupcion-legal-del-embarazo/resource/755e9b54-401d-40a6-8787-c45a8f97937e) from a Mexican public agency about legal pregnancy interruption from 2016 - 2018 (updated in 2023).

```{r}
# loading the dataset
data <- read.csv("ile_2016_2018.csv", stringsAsFactors = F, fileEncoding = "UTF-8")

#Now, create the code to check if the column names are correctly done or not. 
Solution: glimpse(data)
#to check if names are clean and correcting them. 
data_clean_names <-  clean_names(data)
#displaying 
names(data_clean_names)[1:5]
```
Exercise 2 

#dealing with NAs. 

data |> tabyl(numero_hijos) #to check for Nas. 

data |> tabyl(uso_anticonceptivo)
data |> tabyl(uso_anticonceptivo)
data |> tabyl(resultado_ile)

#check for placeholders

data<- data |> convert_to_na(list(numero_hijos = c(-999, ""), uso_anticonceptivo = "-999"))


Exercise 3 

Twoway table: 
table2 <- data_clean_names %>% 
  tabyl(anio, estado_civil) %>% 
  adorn_totals("col") %>%              
  adorn_percentages("row") %>%         
  adorn_pct_formatting(digits = 2) %>% 
  adorn_ns() %>%                      
  adorn_title()


Threeway table: 
data  %>% 
  tabyl(resultado_ile, uso_anticonceptivo, anio) %>% 
  adorn_percentages("all") %>% 
  adorn_pct_formatting(digits = 1) %>%  
  adorn_title() %>% 
  kable() %>% 
  kable_styling(font_size = 8.5)

Exercise 4: Wrap-up 
4.1 
# Sample dataframe with messy data
df <- data.frame(
  'First Name' = c("Alice", "Bob", "Charlie", "David", "Eve", "Frank"),
  'Age ' = c(25, NA, 30, 32, NA, NA),
  'Gender ' = c("Female", "Male", "", "Male", "Female", ""),
  'Alt.Gender' = c("", "", "Male", "", "", "Male"),
  'Comments ' = c("", "", "", "", "", ""),
  'Salary($)' = c(50000, 55000, 60000, 65000, 70000, 75000),
  stringsAsFactors = FALSE
)

4.2 df <- df %>% convert_to_na(list(gender = "", alt_gender = ""))

4.3 df <- df %>% use_first_valid_of(gender, alt_gender)

4.4 df <- df %>% remove_empty("cols")

