---
title: "Introduction to Data Science - Workshop Edition"
subtitle: "Cleaning and wrangling data with `janitor` and `forcats`"
author: "Luis Fernando Ramirez, Elena Dreyer and Shruti Kakade"
institute: "Hertie School | [GRAD-C11/E1339](https://github.com/intro-to-data-science-23/workshop-presentations)" #"`r format(Sys.time(), '%d %B %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Welcome to Janitor! 
In this crash course, we will learn how to clean data in R using a library called Janitor. 

```{r, message = F}
#Preparing the libraries
library(readr)
library(tidyverse)
library(janitor)
library(kableExtra)
```
By using a real-life dataset, we want to demonstrate the perks `janitor` has to offer. First we need to download [this dataset](https://datos.cdmx.gob.mx/dataset/interrupcion-legal-del-embarazo/resource/755e9b54-401d-40a6-8787-c45a8f97937e) from a Mexican public agency about legal pregnancy interruption from 2016 - 2018 (updated in 2023).

```{r}
# loading the dataset
data <- read.csv("ile_2016_2018.csv", stringsAsFactors = F, fileEncoding = "UTF-8")

#Now, create the code to check if the column names are correctly done or not. 

#Now, correct the names of the columns using janitor. 

```

Exercise 2 Dealing with NAs.

```{r}
#In this part of the exercise, it'd be ideal to learn how to deal with NAs.
#Check the following variables and establish a path to deal with NAs. 
#Variables: numero_hijos, uso_anticonceptivo, resultado_ile. 
#For each variable check the amount of NAs, placeholders, and how to deal with them.    

```
Exercise 3: Creating a twoway and threeway table with tabyl and Kable. 
#considering the variables of interest, create a threeway table to relate variables and draw some preliminary conclusions. 

Exercise 4: 






